---
title: "Testing ITPC Data Poolability Across Experimental Conditions"
author: "Analysis Notebook"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    df-print: paged
    embed-resources: true
execute:
  warning: false
  message: false
---

# Overview

This notebook tests whether Inter-Trial Phase Coherence (ITPC) data can be pooled across different experimental conditions in the tDCS-ASSR study.

**Experimental Design:**

- **Period (P):** pre, post (relative to tDCS)
- **Treatment (T):** real, sham (tDCS type)
- **Stimulus State (type):** prestim_itpc, stim_itpc (silence vs 40-Hz stimulus)

**Key Questions:**

1. Is the variance homogeneous across conditions?
2. Are there significant interactions between factors?
3. Can we justify pooling data across certain conditions?

# Setup

```{r setup}
library(tidyverse)
library(brms)
library(bayestestR)
library(ggdist)
library(patchwork)
library(car)        # For Levene's test
library(emmeans)    # For estimated marginal means
library(broom)      # For tidy model outputs
library(GGally)     # For pairs plots

# Set theme
theme_set(theme_minimal(base_size = 12))

# Color palette for conditions
condition_colors <- c(
  "pre_real" = "#0571b0",
  "pre_sham" = "#92c5de",
  "post_real" = "#ca0020",
  "post_sham" = "#f4a582"
)
```

# Data Import

```{r import-data}
# Import ITPC data
itpc_data <- read_csv("../data/Destrieux_final_itpc.csv") %>%
  mutate(
    subject = factor(subject),
    roi = factor(roi),
    task = factor(task),
    P = factor(P, levels = c("pre", "post")),
    T = factor(T, levels = c("real", "sham")),
    type = factor(type, levels = c("prestim_itpc", "stim_itpc")),
    # Create combined condition factor
    condition = interaction(P, T, sep = "_")
  )

# Display data structure
glimpse(itpc_data)

# Summary statistics
summary(itpc_data$value)
```

## Data Overview

```{r data-summary}
# Count observations per condition
itpc_data %>%
  count(P, T, type, name = "n_observations") %>%
  arrange(P, T, type)

# Number of subjects and ROIs
cat("Number of subjects:", n_distinct(itpc_data$subject), "\n")
cat("Number of ROIs:", n_distinct(itpc_data$roi), "\n")
```

# Descriptive Statistics by Condition

```{r descriptive-stats}
# Compute summary statistics by condition
condition_summary <- itpc_data %>%
  group_by(P, T, type, condition) %>%
  summarise(
    n = n(),
    mean = mean(value, na.rm = TRUE),
    median = median(value, na.rm = TRUE),
    sd = sd(value, na.rm = TRUE),
    var = var(value, na.rm = TRUE),
    iqr = IQR(value, na.rm = TRUE),
    min = min(value, na.rm = TRUE),
    max = max(value, na.rm = TRUE),
    .groups = "drop"
  )

condition_summary
```

# Visual Inspection of Distributions

## Overall Distribution by Condition

```{r dist-by-condition, fig.width=10, fig.height=6}
ggplot(itpc_data, aes(x = value, fill = condition)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~type, ncol = 1) +
  scale_fill_manual(values = condition_colors) +
  labs(
    title = "ITPC Distribution by Experimental Condition",
    x = "Z-normalized ITPC",
    y = "Density",
    fill = "Condition"
  ) +
  theme(legend.position = "bottom")
```

## Boxplots by Condition

```{r boxplots-condition, fig.width=10, fig.height=6}
ggplot(itpc_data, aes(x = condition, y = value, fill = condition)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  facet_wrap(~type, ncol = 1, scales = "free_y") +
  scale_fill_manual(values = condition_colors) +
  labs(
    title = "ITPC Values by Experimental Condition",
    x = "Condition",
    y = "Z-normalized ITPC"
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

## Violin Plots with Summary Statistics

```{r violin-plots, fig.width=10, fig.height=8}
ggplot(itpc_data, aes(x = P, y = value, fill = T)) +
  geom_violin(alpha = 0.6, draw_quantiles = c(0.25, 0.5, 0.75)) +
  geom_boxplot(width = 0.2, alpha = 0.8, position = position_dodge(0.9)) +
  facet_wrap(~type, ncol = 1) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "ITPC Distribution: Period × Treatment Interaction",
    x = "Period",
    y = "Z-normalized ITPC",
    fill = "Treatment"
  ) +
  theme(legend.position = "bottom")
```

# Statistical Tests for Homogeneity

## Levene's Test for Homogeneity of Variance

```{r levene-test}
# Levene's test by stimulus type
levene_prestim <- itpc_data %>%
  filter(type == "prestim_itpc") %>%
  leveneTest(value ~ condition, data = .)

levene_stim <- itpc_data %>%
  filter(type == "stim_itpc") %>%
  leveneTest(value ~ condition, data = .)

cat("Levene's Test for Prestim ITPC:\n")
print(levene_prestim)

cat("\nLevene's Test for Stim ITPC:\n")
print(levene_stim)

# Interpretation
cat("\nInterpretation:\n")
cat("If p < 0.05: Variances are significantly different (pooling questionable)\n")
cat("If p >= 0.05: Variances are homogeneous (pooling may be justified)\n")
```

## Bartlett's Test (parametric alternative)

```{r bartlett-test}
# Bartlett's test
bartlett_prestim <- itpc_data %>%
  filter(type == "prestim_itpc") %>%
  bartlett.test(value ~ condition, data = .)

bartlett_stim <- itpc_data %>%
  filter(type == "stim_itpc") %>%
  bartlett.test(value ~ condition, data = .)

cat("Bartlett's Test for Prestim ITPC:\n")
print(bartlett_prestim)

cat("\nBartlett's Test for Stim ITPC:\n")
print(bartlett_stim)
```

## Variance Ratio Analysis

```{r variance-ratios}
# Compute variance ratios across conditions
variance_summary <- condition_summary %>%
  group_by(type) %>%
  summarise(
    max_var = max(var),
    min_var = min(var),
    variance_ratio = max_var / min_var,
    .groups = "drop"
  )

cat("Variance Ratios by Stimulus Type:\n")
print(variance_summary)
cat("\nRule of thumb: Variance ratio < 3 suggests reasonable homogeneity\n")
```

# Mixed Effects Model Approach

## Fit Full Factorial Model

```{r fit-lmer-model}
library(lme4)
library(lmerTest)

# Fit mixed effects model with all interactions
# Random effects: subject and ROI
model_full <- itpc_data %>%
  lmer(value ~ P * T * type + (1 | subject) + (1 | roi), data = .)

# Model summary
summary(model_full)
```

## ANOVA Table

```{r anova-table}
# Type III ANOVA
anova(model_full, type = "III")
```

## Estimated Marginal Means

```{r emmeans-analysis}
# Estimated marginal means for each condition
emm_condition <- emmeans(model_full, ~ P * T * type)
print(emm_condition)

# Pairwise comparisons
cat("\nPairwise Comparisons (P × T within each stimulus type):\n")
pairs(emmeans(model_full, ~ P * T | type), adjust = "tukey")
```

# Interaction Plots

```{r interaction-plots, fig.width=12, fig.height=6}
# Extract estimated marginal means
emm_df <- as.data.frame(emm_condition)

# Plot 1: P × T interaction by stimulus type
p1 <- ggplot(emm_df, aes(x = P, y = emmean, color = T, group = T)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.1) +
  facet_wrap(~type) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Period × Treatment Interaction",
    x = "Period",
    y = "Estimated Marginal Mean (ITPC)",
    color = "Treatment"
  ) +
  theme(legend.position = "bottom")

# Plot 2: P × type interaction by treatment
p2 <- ggplot(emm_df, aes(x = P, y = emmean, color = type, group = type)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.1) +
  facet_wrap(~T) +
  scale_color_brewer(palette = "Dark2") +
  labs(
    title = "Period × Stimulus Type Interaction",
    x = "Period",
    y = "Estimated Marginal Mean (ITPC)",
    color = "Stimulus Type"
  ) +
  theme(legend.position = "bottom")

p1 
```

# Intraclass Correlation Coefficients

```{r icc-analysis}
library(performance)

# ICC for subject and ROI
icc_results <- icc(model_full)
print(icc_results)

cat("\nInterpretation:\n")
cat("ICC close to 0: Little clustering (pooling across subjects/ROIs justified)\n")
cat("ICC close to 1: Strong clustering (need hierarchical modeling)\n")
```

# Bayesian Approach to Heterogeneity

## Fit Bayesian Model with Varying Intercepts

```{r fit-brms-model, cache=TRUE, results='hide'}
# Fit Bayesian mixed effects model
# This allows us to quantify uncertainty in variance components
bayes_model <- brm(
  value ~ P * T * type + (1 | subject) + (1 | roi),
  data = itpc_data,
  family = gaussian(),
  prior = c(
    prior(normal(0, 2), class = Intercept),
    prior(normal(0, 1), class = b),
    prior(student_t(3, 0, 2.5), class = sd),
    prior(student_t(3, 0, 2.5), class = sigma)
  ),
  chains = 4,
  iter = 2000,
  warmup = 1000,
  cores = 4,
  seed = 123,
  backend = "cmdstanr",
  silent = 2,
  refresh = 0
)
```

```{r bayes-summary}
# Model summary
summary(bayes_model)
```

## Posterior Predictive Checks

```{r pp-checks, fig.width=10, fig.height=6}
# Posterior predictive checks
pp_check(bayes_model, ndraws = 100) +
  labs(title = "Posterior Predictive Check")

# By condition
pp_check(bayes_model, type = "stat_grouped", group = "condition", stat = "mean") +
  labs(title = "Posterior Predictive Check: Mean by Condition")
```

## Extract Fixed Effects Posteriors

```{r bayes-effects}
# Extract posterior samples for fixed effects
fixed_effects <- bayes_model %>%
  as_draws_df() %>%
  select(starts_with("b_")) %>%
  pivot_longer(everything(), names_to = "parameter", values_to = "value")

# Plot posterior distributions of interaction effects
ggplot(fixed_effects %>% filter(str_detect(parameter, ":")),
       aes(x = value, y = parameter)) +
  stat_halfeye(.width = c(0.8, 0.95)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Posterior Distributions of Interaction Effects",
    x = "Effect Size",
    y = "Parameter"
  )
```

# Test Specific Pooling Hypotheses

## Can we pool across Period (pre/post)?

```{r test-period-pooling}
# Test if Period main effect and interactions are negligible
period_tests <- hypothesis(
  bayes_model,
  c("Ppost = 0",
    "Ppost:Tsham = 0",
    "Ppost:typestim_itpc = 0",
    "Ppost:Tsham:typestim_itpc = 0")
)

print(period_tests)

cat("\nInterpretation:\n")
cat("Evidence Ratio > 3: Moderate evidence AGAINST pooling\n")
cat("Evidence Ratio < 1/3: Moderate evidence FOR pooling\n")
```

## Can we pool across Treatment (real/sham)?

```{r test-treatment-pooling}
# Test if Treatment main effect and interactions are negligible
treatment_tests <- hypothesis(
  bayes_model,
  c("Tsham = 0",
    "Ppost:Tsham = 0",
    "Tsham:typestim_itpc = 0",
    "Ppost:Tsham:typestim_itpc = 0")
)

print(treatment_tests)
```

## Can we pool across Stimulus Type (prestim/stim)?

```{r test-type-pooling}
# Test if Stimulus Type main effect and interactions are negligible
type_tests <- hypothesis(
  bayes_model,
  c("typestim_itpc = 0",
    "Ppost:typestim_itpc = 0",
    "Tsham:typestim_itpc = 0",
    "Ppost:Tsham:typestim_itpc = 0")
)

print(type_tests)
```

# Subject-Level Heterogeneity Analysis

```{r subject-heterogeneity, fig.width=12, fig.height=8}
# Compute subject-level means by condition
subject_means <- itpc_data %>%
  group_by(subject, P, T, type) %>%
  summarise(mean_itpc = mean(value, na.rm = TRUE), .groups = "drop")

# Plot subject trajectories
ggplot(subject_means, aes(x = P, y = mean_itpc, group = subject, color = subject)) +
  geom_line(alpha = 0.6) +
  geom_point(alpha = 0.8) +
  facet_grid(type ~ T) +
  labs(
    title = "Individual Subject Trajectories by Condition",
    x = "Period",
    y = "Mean ITPC",
    color = "Subject"
  ) +
  theme(legend.position = "right")
```

# ROI-Level Heterogeneity Analysis

```{r roi-heterogeneity, fig.width=10, fig.height=6}
# Compute variance of ITPC across conditions for each ROI
roi_variance <- itpc_data %>%
  group_by(roi, type) %>%
  summarise(
    var_itpc = var(value, na.rm = TRUE),
    mean_itpc = mean(value, na.rm = TRUE),
    .groups = "drop"
  )

# Plot variance distribution across ROIs
ggplot(roi_variance, aes(x = var_itpc, fill = type)) +
  geom_histogram(bins = 30, alpha = 0.7, position = "identity") +
  facet_wrap(~type, ncol = 1) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Distribution of ITPC Variance Across ROIs",
    x = "Variance",
    y = "Count",
    fill = "Stimulus Type"
  )
```

# Correlation Analysis Between Conditions

```{r condition-correlations, fig.width=10, fig.height=10}
# Create wide format for correlation analysis
itpc_wide <- itpc_data %>%
  group_by(subject, roi, P, T, type) %>%
  summarise(mean_itpc = mean(value, na.rm = TRUE), .groups = "drop") %>%
  unite("condition_type", P, T, type, sep = "_") %>%
  pivot_wider(names_from = condition_type, values_from = mean_itpc)

# Compute correlations
cor_matrix <- itpc_wide %>%
  select(-subject, -roi) %>%
  cor(use = "pairwise.complete.obs")

# Heatmap
library(corrplot)
corrplot(cor_matrix, method = "color", type = "upper",
         tl.col = "black", tl.srt = 45,
         addCoef.col = "black", number.cex = 0.7,
         title = "Correlation Matrix: ITPC Across Conditions",
         mar = c(0, 0, 2, 0))
```

# Summary and Recommendations

```{r summary-report}
cat("=" %>% str_dup(70), "\n")
cat("SUMMARY: ITPC DATA POOLABILITY ANALYSIS\n")
cat("=" %>% str_dup(70), "\n\n")

cat("1. VARIANCE HOMOGENEITY:\n")
cat("   - Check Levene's and Bartlett's test p-values above\n")
cat("   - Check variance ratios (should be < 3 for pooling)\n\n")

cat("2. INTERACTION EFFECTS:\n")
cat("   - Significant interactions suggest conditions differ\n")
cat("   - Check ANOVA table and Bayesian hypothesis tests\n\n")

cat("3. HIERARCHICAL STRUCTURE:\n")
cat("   - ICC values indicate clustering by subject and ROI\n")
cat("   - Mixed models account for this structure\n\n")

cat("4. RECOMMENDATIONS:\n")
cat("   Based on the analyses above:\n")
cat("   \n")
cat("   a) If interactions are significant:\n")
cat("      → DO NOT pool across those factors\n")
cat("      → Analyze conditions separately or model interactions\n")
cat("   \n")
cat("   b) If variances differ substantially:\n")
cat("      → Consider variance-stabilizing transformations\n")
cat("      → Use robust statistical methods\n")
cat("   \n")
cat("   c) Given the hierarchical structure:\n")
cat("      → Always use mixed models with random effects\n")
cat("      → Account for subject and ROI clustering\n")
cat("   \n")
cat("   d) Stimulus type (prestim vs stim) typically shows:\n")
cat("      → Strong differences (fundamental to the design)\n")
cat("      → Should NOT be pooled\n")
cat("=" %>% str_dup(70), "\n")
```

# Session Info

```{r session-info}
sessionInfo()
```
